---
title: "My Data"
author: "Ilkka Kiistala"
date: "7 Apr 2015"
output:
  pdf_document:
    number_sections: yes
---

# Baana Biker data

I'm going to combine biker data collected at Baana and Helsinki weather data.

## Load biker data

Load biker data from Helsinki Region Infoshare:

```{r, cache=TRUE}
url <- 'http://www.hel.fi/hel2/tietokeskus/data/helsinki/ksv/Baanan_pyorailijamaarat.xlsx'
download.file(url, destfile="bikers.xlsx")

library(xlsx)
bikers <- read.xlsx("bikers.xlsx", 1)
```

## Check biker data

```{r}
head(bikers)
nrow(bikers)
str(bikers)
```

## Cleanup

We only need date and number of bikers.

```{r}
bikers <- bikers[, c(2,4)]
```

Let's rename the columns.

```{r}
head(bikers)
names(bikers) <- c("date", "bikers")
head(bikers)
```

---

# Load weather data

## Helsinki daily temperature observations

The Finnish Meteorological Institute (FMI) is a research and service agency under the Ministry of Transport and Communications.

Observations are accessible via FMI Open Data WFS service. Quering the service requires registration, which provides user with an API key.

```{r eval=FALSE}
# Example API call:
http://data.fmi.fi/fmi-apikey/insert-your-apikey-here/wfs?request=getFeature&storedquery_id=fmi::observations::weather::daily::simple&place=helsinki
```
See their Open Data Manual: http://en.ilmatieteenlaitos.fi/open-data-manual-fmi-wfs-services

## Pre-editing the data

After fetching the data as XML, it was parsed into following form:

```{r eval=FALSE}
2013-01-01 tday 2.8
2013-01-02 tday 2.3
2013-01-03 tday 1.0
2013-01-04 tday 1.6
2013-01-05 tday -2.5
2013-01-06 tday -4.6
2013-01-07 tday -4.7
2013-01-08 tday -0.1
```

## Importing the temperature data into R

```{r}
tem <- read.csv("helsinki-temperatures.tsv", sep=" ", header=FALSE, stringsAsFactors=FALSE)
summary(tem)
```
Summary reveals some repeating days, so to be sure, we need to check that their data match.
But before that, let's remove the columns with 'tday' values and name our columns.

```{r}
names(tem)
tem[,2] <- NULL
names(tem)

names(tem) <- c("date", "tday")
head(tem)
```

Now it's easier to refer to the columns. Let's check those repeating dates.

```{r}
date_count <- as.data.frame(table(tem$date), stringsAsFactors=FALSE)
# subset(date_count, Freq > 1)
# multidate <- as.vector(subset(date_count, Freq > 1, select="Var1"))
multidate <- subset(date_count, Freq > 1, select="Var1")[,1]

# tem[tem$date %in% as.vector(multidate),]
head( tem[tem$date %in% multidate,] )

```

No anomalies in the pairs.
It is safe to remove duplicates.

```{r}
nrow(tem)
tail(tem)
tem <- tem[!duplicated(tem),]
nrow(tem)
tail(tem)
head( tem[tem$date %in% multidate,] )

# This recalculates rownames
rownames(tem) <- NULL
tail(tem)
```

# Merging the two data frames

```{r}
nrow(bikers)
nrow(tem)
```

Let's find out the latest dates we have in the two data sets.

```{r}
max(bikers$date)
max(tem$date)

# the earlier date is usually in the biker data, but let's play it safe
mutually_latest_date <- min(max(bikers$date), max(tem$date))
mutually_latest_date

# check that row count matches
nrow(subset(bikers, date <= mutually_latest_date))
nrow(subset(tem, date <= mutually_latest_date))

bikers <- subset(bikers, date <= mutually_latest_date)
tem <- subset(tem, date <= mutually_latest_date)

```

Now their row counts match, so let's merge them.
First, some values to check after merge:


```{r}
tail(bikers);tail(tem)
```

Merge.

```{r}
# bw as in "Bikers and Weather"
bw <- merge(bikers, tem)
nrow(bw)
```

The date column is of different type.
We need to transform tem$date into Date type.

```{r}
tem$date <- as.Date(tem$date)
bw <- merge(bikers, tem)
nrow(bw)
tail(bw)
```

## Add weekday data column

```{r}
bw$weekday <- format( as.Date(bw$date), "%w" )
```

# Description of columns

* date: the date of the biker count / weather observations
* bikers: number of bikers passing the electronic counter device at Baana
* tday: average temperature of the day  
* weekday: weekday of the day, 0=Sunday, 1=Monday  
