---
title: "Plots"
author: "Pedro J. Aphalo"
date: "31 March 2015"
output:
  beamer_presentation:
    fig_height: 4
    fig_width: 6
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align='center', size="tiny",
               fig.width=6, fig.height=4, 
               out.width='.75\\textwidth')
```

## Slide with R Code and Output

```{r}
summary(cars)
```

## Slide with Plot

A very simple plot using example data included in R and base R plot functions

```{r}
plot(cars)
```

## ggplot2

Nowadays the package most frequently used for _normal_ plotting is `ggplot2`

- it is based on a grammar of graphics
- in plain words, we assemble plots by adding different _pieces_
- these pieces are quite modular
- this allows making many different types of plots from not so many _pieces_
- the basic pieces are _layers_
- the name reflects the fact that order in which we add them is significant
- later additions are plotted on top of the first ones

## what is a ggplot object?

- a `ggplot` object is independent of the output format
- it is not purely visual, it is to a significant extent semantic
- the size and the `theme` can be decided at the time of generating output
- a `ggplot` can be modified by adding and sometimes replacing elements
- a `ggplot` object can be large (in memory) because it contains a copy of the data plotted
- this means that the `ggplot` object is self-sufficient (you can copy it to a new 'empty' R session and still generate the output)

## What are the words of the grammar
- data
- aesthetics
- statistics
- scales
- geometries

## data
- `data` tells R where to look for the variables used in the plot, data is supplied
as a `data.frame` object
-  _aesthetics_ (`aes`) tells how to map variables to features in the plot
   * stablishes a connection between a variable an aesthetic _dimension_
   * for example: `x` coordinate, `colour`, etc.
-  _statistics_ or some operation on the data before plotting
_  _scales_ the mapping between data values and aesthetic values
   * e.g. is value 1 by plotted as red and value 2 with say blue, or using some other colours
   
## geometries
- they are easier to describe with examples
- `geom_line`, `geom_point`, `geom_text` 
- each geom you add to a ggplot adds a new layer to the plot (on top of the layers added earlier)

## annotations
- they add layers that are not directly 'connected' to the data
- e.g. a label not directly related to a single observation
- annotations can also use different geoms and aesthetics, buy they are not inherited, they are
selected locally to each annotation

## A very simple plot
```{r}
library(ggplot2)
# runif generates random numbers
# we create a data frame to play with
my.df <- data.frame(xx = 1:10, yy = runif(10))
# we create a plot and save it as 'my.plot'
my.plot <- ggplot(data = my.df, aes(x = xx, y = yy)) + 
  geom_point()
```

## _Printing_ the plot
```{r}
print(my.plot)
```

## _Printing_ the plot using a different `theme`
```{r}
print(my.plot + theme_bw())
```

## _Printing_ the plot using a different base font size
```{r}
print(my.plot + theme_bw(15))
```

## What is in my.plot

```{r}
str(my.plot)
```

## A not so simple plot
```{r}
# runif generates random numbers
# we create a data frame to play with
my.2nd.df <- 
  data.frame(xx = rep(1:10, 2), 
             yy = runif(20), 
             group = factor(rep(c("A", "B"), c(10, 10))) )
# we create a plot and save it as 'my.plot'
my.2nd.plot <- ggplot(data = my.2nd.df,
                      aes(x = xx, y = yy, colour = group)) + 
  geom_point()
```

## _Printing_ the plot
```{r}
print(my.2nd.plot)
```


## A not so simple plot
```{r}
# runif generates random numbers
# we create a data frame to play with
my.2nd.df <- 
  data.frame(xx = rep(1:10, 2), 
             yy = runif(20), 
             group = factor(rep(c("A", "B"), c(10, 10))) )
# we create a plot and save it as 'my.plot'
my.3rd.plot <- ggplot(data = my.2nd.df,
                      aes(x = xx, y = yy, shape = group)) + 
  geom_point()
```

## _Printing_ the plot
```{r}
print(my.3rd.plot)
```

## A plot with two aesthetics for `group`
```{r}
# we create a plot and save it as 'my.plot'
my.4th.plot <- ggplot(data = my.2nd.df,
                      aes(x = xx, y = yy, 
                          shape = group,
                          colour = group)) + 
  geom_point()
```

## _Printing_ the plot
```{r}
print(my.4th.plot)
```

## Replacing the default labels
```{r}
my.5th.plot <- my.4th.plot +
  labs(x="Time (d)", y = "Response (a.u)", 
       colour = "Treatment", shape = "Treatment",
       title = "Artificial data")
print(my.5th.plot)
```

## Adding another layer on the fly
```{r}
print(my.5th.plot + geom_line())
```

## Adding another layer on the fly
```{r}
print(my.5th.plot + geom_smooth())
```

## Another way of plotting the same data
```{r}
# we create a plot and save it as 'my.plot'
my.6th.plot <- ggplot(data = my.2nd.df,
                      aes(x = group, y = yy))
print(my.6th.plot + geom_boxplot())
```

## Another way of plotting the same data
```{r}
print(my.6th.plot + geom_violin())
```

## A larger set of normally distributed data
```{r}
# rnorm generates random numbers
my.3rd.df <- 
  data.frame(yy = c(rnorm(n = 50, mean = 2), rnorm(n = 50, mean = 0)),
             group = factor(rep(c("A", "B"), c(50, 50))) )
my.7th.plot <- ggplot(data = my.3rd.df,
                      aes(x = group, y = yy))
print(my.7th.plot + geom_boxplot())
```

## The plot with a few additional layers

```{r}
print(my.7th.plot + 
        geom_violin() +
        geom_point(aes(colour = group)) +
        geom_rug(aes(colour = group), 
                 side = "l", alpha = 0.5) )
```

