---
title: "Script example (R Markdwon)"
author: "Pedro J. Aphalo"
date: "30 March 2015"
output:
  pdf_document:
    number_sections: yes
    toc: yes
---

# knitr setup

```{r}
library(knitr)
```

# Data

We use cork production data from package `agridat`.

```{r}
library(agridat)
data("box.cork")

summary(box.cork)
```

# Data preparation

We melt from 4 columns into a single column.

```{r}
library(reshape2)
cork.melted <- melt(box.cork, 
                    id.vars = "tree", 
                    variable.name = "trunk.side", 
                    value.name = "cork.cg")
```

# EDA (exploratory data analysis)}

We do exploratory plotting with `ggplot2'.

```{r}
library(ggplot2)
```


First you save a simple *dot plot.*

```{r}
fig.cork <- ggplot(data = cork.melted, 
                   aes(x = trunk.side, y = cork.cg)) + 
  geom_point()
fig.cork
```

We add to the plot saved above *boxplots*, and also change
the $y$ scale to logarithmic.

<<>>=
fig.cork + geom_boxplot(fill = NA)
fig.cork + geom_boxplot(fill = NA) + scale_y_log10()
@

We add to the plot saved above \emph{violin plots}, and also change
the $y$ scale to logarithmic.

```{r}
fig.cork + geom_violin(fill = NA)
fig.cork + geom_violin(fill = NA) + scale_y_log10()
```

We create and save a new plot, with fitted densities to
assess the shape of the distributions and their differences.

```{r}
fig.cork1 <- ggplot(data = cork.melted, 
                    aes(x = cork.cg, colour = trunk.side)) + 
  geom_density() + xlim(0, NA)
fig.cork1
```


Here we use histograms instaed.

```{r}
fig.cork2 <- ggplot(data = cork.melted, aes(x = cork.cg, fill = trunk.side)) + 
  geom_histogram(binwidth = 10) + 
  facet_wrap(~trunk.side, ncol = 2)
fig.cork2
```

# Test of significance

We fit a linear model, calculate and ANOVA table
and check diagnosis plots

```{r}
cork.fit <- lm(log10(cork.cg) ~ trunk.side + tree, data = cork.melted)
anova(cork.fit)
plot(cork.fit)
```
